Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-01 image_analysis_paper_development2/SI_text.tex
updated paper

Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between time image pixels is informative.
%
We want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
%
We therefore must adjust images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.

The relevant image operations are 
\begin{enumerate}
\item {\bf Intensity normalization:} Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
%
\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. 
%
\item {\bf Intensity scaling (multichannel images only):} The \texttt{immultiply} function is used to scale signal intensities. This is important for multichannel images, and it determines the (relative) contributions of the various signals.  
%
\item {\bf Mean-centering:}  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function). The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
%
\item {\bf Size scaling:} For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
%
\item {\bf Removing corners:} The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity). This removes any corner effects under rotations. 
\end{enumerate}

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon^2$ to be $1/10^{th}$ of the median of the squared pairwise distances.

@@

To register sets of images, 
the first step is to compute the optimal alignments between pairs of images.
%
Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
%
For each image pair $I_i$ and $I_j$ we compute
\begin{equation}\label{eq:opt_pairwise}
\theta_{ij} = \argmin_{
0^\circ \le \theta < 360^\circ }
 \|g(I_j, \theta) - I_i \|^2.
\end{equation}
where $g(I_j, \theta)$ is image $I_j$ rotated around the center of the square by $\theta$ degrees.
%
The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels.
%
The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
%
Image rotation is performed with the \texttt{imwarp} function in Matlab, using linear interpolation to estimate the pixel intensities after rotation.
%
The missing pixels in the corners of the rotated image are taken to have zero intensity.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution (for the results presened, we use $10^\circ$ discretization steps).
%
%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
%
Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
%
Once we have computed $\theta_{ij}$ for all image pairs, we can proceed with the vector diffusion maps procedure.
%
The rotation matrices returned from vector diffusion maps can be used to calculate the angle of rotation for each image (by inverting \eqref{eq:R_theta}), and the function $g$ as described previously is then used to rotate the images.

@@

Once we have registered and ordered the images, we can smooth the resulting trajectory to obtain a ``stereotypic" developmental trajectory.
%
Let $I_1, \dots, I_m$ denote the set of {\em registered and ordered} images (so $I_j$ is the $j^{th}$ image in the ordered set).
%
We define the average image at time $\tau$, denoted $\overline{I}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{I}_{\tau} =  \frac{\sum_j \exp \left( - \frac{|j - \tau|^2}{\sigma^2} \right) I_i}{\sum_j \exp \left( - \frac{|j - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter.
%
For the images in \fig~4, we take $\sigma = 3$.
%
See \citep{kemelmacher2011exploring} for a more detailed discussion.

>>>>

Before applying vector diffusion maps, images must be preprocessed so that the Euclidean distance between the image pixels is informative: we want images who are developmentally similar to have relatively small Euclidean distances, and images who are at disparate developmental time points to have relatively large distances. 
%
We must preprocess images to remove any experimental and imaging artifacts, so that most of the variation is relevant to the developmental dynamics.

For our general purposes, the relevant image operations are 
\begin{enumerate}
\item {\bf Intensity normalization:} Contrast-limited adaptive histogram equalization (using the \texttt{adapthisteq} function with an $8 \times 8$ tile grid, and a uniform distribution for the intensities with a clip limit of 0.01) is used to normalize the intensities of signals whose absolute intensity is not meaningful/informative.
%
\item {\bf Blur:} The \texttt{imfilter} function with a disc filter is used to blur signals whose small-scale structure is not informative. For example, 
%
\item {\bf Intensity scaling (multichannel images only):} The \texttt{immultiply} function is used to scale signal intensities. This is important for multichannel images, and it determines the (relative) contributions of the various signals.  
%
\item {\bf Mean-centering:}  The Canny method \citep{canny1986computational} is used to detect the edges of the object in each image (using the \texttt{edge} function). The image is then translated so that the object (as determined by extremeties of the detected edges) is centered.
%
\item {\bf Size scaling:} For images whose relative size is unimportant to the developmental dynamics, the images are rescaled/dialated so that the object size (as determined by extremeties of the detected edges) is constant (we set this to be 80\% of the total image).
%
\item {\bf Removing corners:} The image is cropped/filtered using a disc centered in the middle of the image and whose diameter is equal to the number of pixels; any pixels outside of this disc are set to 0 (no intensity). This removes any corner effects under rotations. 
\end{enumerate}

@@

Given $m$ data points $x_1, \dots, x_m$ (typically vectors in a high-dimensional vector space), we want to find a coordinate transformation $y(x)$ that preserves local geometry: points that are ``close" in the original space should also be ``close" in the coordinates $y$.
%
The first step is to construct the matrix $W \in \mathbb{R}^{m \times m}$, where $W_{ij}$ is large if points $x_i$ and $x_j$ are ``close.''
%
We use a diffusion kernel,
\begin{equation} \label{eq:dmaps_W}
W_{ij} = \exp \left( -\frac{d^2(x_i, x_j)}{\epsilon^2} \right),
%W_{ij} = e^{ -\frac{d^2(x_i, x_j)}{\epsilon^2}}
\end{equation}
where $d(x_i, x_j)$ is a pairwise distance between $x_i$ and $x_j$ (often the Euclidean distance), and $\epsilon$ is a characteristic scale.
%
Points less than $\epsilon$ apart are thus considered ``close'' and points farther than $\epsilon$ apart are considered ``far away''.
%
$\epsilon$ can be chosen using several techniques (see, for example \citep{coifman2008graph, rohrdanz2011determination}); here, we take $\epsilon$ to be $1/4-1/2$ of the median of the pairwise distances.

@@

To register sets of images, 
the first step is to compute the optimal alignments between pairs of images.
%
Practically, we have square images discretized as pixels (rather than continuous functions on the plane).
%
For each image pair $I_i$ and $I_j$ we compute
\begin{equation}\label{eq:opt_pairwise}
\theta_{ij} = \argmin_{
0^\circ \le \theta < 360^\circ }
 \|g(I_j, \theta) - I_i \|^2.
\end{equation}
where $g(I_j, \theta)$ is image $I_j$ rotated around the center of the square by $\theta$ degrees.
%
The norm, $\| \cdot \|$, is the Euclidean norm between the pixel intensities of the channels.
%
The domain of the image (a square) is not invariant to our rotations; however, the pixels near the corners of the square are preprocessed to have zero intensity, and so the norm can be meaningfully computed as long as the main image does not ``move out of'' the original square.
%
Image rotation is performed with the \texttt{imwarp} function in Matlab, using linear interpolation to estimate the pixel intensities after rotation.
%
The missing pixels in the corners of the rotated image are taken to have zero intensity.
%
The solution to \eqref{eq:opt_pairwise} is not easily computed, as the objective function will most likely be nonconvex.
%
Therefore, instead of using an optimization procedure, we discretize the search space and exhaustively search to find the solution (for the results presened, we use $10^\circ$ discretization steps).
%
%We discritize the search space of rotations into 20 possible rotations %($d\theta  \in \{0, \pi/10, \pi/5, \dots, 9 \pi/5, 19\pi/10 \}$),
%and 11 possible translations in both . %($dx, dy \in \{-20, -16, -12, \dots, 12, 16, 20 \}$).
%
%We check all possible combinations for the rotation and translations that best align $I_j$ to $I_i$.
%
Although computationally demanding, this ``embarrassingly parallelizable" direct enumeration approach is not prohibitive here.
%
Once we have computed $\theta_{ij}$ for all image pairs, we can proceed with the vector diffusion maps procedure.
%
The rotation matrices returned from vector diffusion maps can be used to calculate the angle of rotation for each image (by inverting \eqref{eq:R_theta}), and the function $g$ as described previously is then used to rotate the images.

@@

Once we have registered and ordered the images, we can smooth the resulting trajectory to obtain a ``stereotypic" developmental trajectory.
%
Let $I_1, \dots, I_m$ denote the set of {\em registered and ordered} images (so $I_j$ is the $j^{th}$ image in the ordered set).
%
We define the average image at time $\tau$, denoted $\overline{I}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{I}_{\tau} =  \frac{\sum_j \exp \left( - \frac{|j - \tau|^2}{\sigma^2} \right) I_i}{\sum_j \exp \left( - \frac{|j - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter.
%
For the images in \fig~5~and~6, we take $\sigma = 2$.
%
See \citep{kemelmacher2011exploring} for a more detailed discussion.

### 2015-02-01 image_analysis_paper_development2/code_for_distribution/README/README.tex
updated paper

@@

The basic steps are as follows:
\begin{enumerate}
   \item Read in images
   \item Preprocess the images (smooth and/or equalize channels, mean-center, etc.)
   \item Run vector diffusion maps to register and order preprocessed images
   \item (Optional) Save registered and ordered images
\end{enumerate}

@@

The fields are as follows
\begin{description}
%
\item[2D/3D] Select ``2D'' if your images are traditional 2D images. Select ``3D'' if your images are z-stacks.
%
\item[Image Directory] Enter the directory where your images or z-stacks are stored. This directory can be an absolute path (e.g. \texttt{C:/Documents/Images}) or a path relative to the directory where this code is stored (e.g. \texttt{../images})
%
\item[Image Prefix] The prefix for all image file names. The images are assumed to named as the image prefix, followed by a number. 
%
\item[Stack Prefix] (only for 3D z-stacks) The prefix for all stack folder names. Each stack is assumed to be stored in a single folder, named as the stack prefix, followed by a number. 
%
\item[Image Extension] The extension of the images (e.g., tif, jpg)
%
\item[Number of Images] The number of images or z-stacks. Images are assumed to be stored as the image prefix, followed by a two-digit index, where the indices run from 1 to the number of images (e.g., image\_prefix01.tif, image\_prefix02.tif, etc.). For three-dimensional images, the z-stacks are assumed to be stored in individual folders whose names are the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of stacks (e.g., z\_stack\_prefix01, z\_stack\_prefix02, etc.)
%
\item[Number of Images in Stack] (only for z-stacks) The number of images per z-stack. Images are assumed to be stored in the individual stack directories as the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of images in the stack (e.g., stack\_prefix01.tif, stack\_prefix02.tif, etc.). 
%
\end{description}

After populating the required fields, select the ``Read Images'' button to read in the images (a wait par will appear to show the progress of image reading).
%
After reading the images, select the ``Show Images'' button to display the images (z-stacks will be displayed as maximum intensity projections). This step is optional.

@@

\begin{description}
\item[Weight] Each channel in the images is scaled by the weight factor, where all weights at 1 does not adjust the channel intensities. For grayscale images, only the first weight is used. 
%
\item[Blur] Each channel in the images is blurred using a disc factor with radius equal to a fraction of the image as indicated by the ``blur'' field. For example, setting the blur to 5\% means that the image is blurred using a disc filter with radius $0.5 \times$ the number of pixels. 
%
\item[Normalize] If the normalize option is selected for a specific channel, then Contrast-Limited Adaptive Histogram Equalization is applied to that channel to normalize the intensities across the image. This is important for signals whose absolute intensity is not meaningful or informative. 
%
\item[Use to Mean-Center] If the mean-center option is selected for any of the channels, then the object is centered in the image by detecting the edges of the object given by the selected channels. If no mean-center options are selected, then the object is not mean-centered.
%
\item[Resize Images] If the resize images option is selected, then each image is scaled/dialated so that the object occupies 80\% of the total image frame. This removes any size effects/variations in the images. 
%
\item[Number of Pixels] The number of pixels wanted in the images for analysis. This does {\em not} have to be the same number of pixels as the true resolution of the images. The images are assumed to be square. 
%
\end{description}

After populating the required fields, select the ``Apply Image'' button to apply the selected preprocessing options (a wait bar will appear to show the progress).
%
After preprocessing the images, select the ``Show Images'' button to display the preprocessed images (z-stacks will be displayed as maximum intensity projections). This step is optional.

@@

After preprocessing the images, one can apply various analysis algorithms. 
%
There are three parameters
%
\begin{description}
%
\item[Angular Discretization (degrees)] (required for image registration) This is the angular discretization used to search for the pairwise alignments between images. For example, if the angular discretization is 10$^\circ$, then the pairwise alignments are computed by computing the image discrepancy after 10$^\circ$ rotation, after a 20$^\circ$ rotation, ..., up to a 360$^\circ$ rotation, and then selecting the rotation which minimizes the discrepancy. Lower values of the angular discretization increase the accuracy of the method, but also increase the computational cost. 
%
\item[Kernel Scale] This is the scaling of the diffusion maps kernel. This parameter should be $\mathcal{O}(1)$, and we have empirically found for our data sets that values of $0.25-0.5$ often yield good results. 
%
\end{description}

After this, select the ``Apply to Raw Images'' option to apply the computed registration and/or ordering to the original images; select the ``Apply to Preprocessed Images'' option to apply the computed registration and/or ordering to the preprocessed images (after blurring, normalization, etc.)

There are four analysis options
\begin{description}
%
\item[Register] (this option is only for two-dimensional images) This registers the imaging data set using vector diffusion maps to compute the optimal rotations. 
%
\item[Order] This orders the imaging data set using diffusion maps. {\em Important:} This option assumes that the imaging data set is preregistered! This option can be used for imaging data sets that were preregistered using some {\em a priori} knowledge about the system. 
%
\item[Register + Order] (this option is only for two-dimensional images) This registers and orders the imaging data set using vector diffusion maps to compute the optimal rotations and the temporal ordering. 
%
\item[Register + Order z-stacks] (this option is only for three-dimensional z-stacks) This registers and orders the imaging data set. It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections using vector diffusion maps.  It then orders the data using diffusion maps on the (now registered) three-dimensional z-stacks. 
%
\end{description}

After analysis select the ``Show Images'' button to display the registered and/or images (z-stacks will be displayed as maximum intensity projections). This step is optional.

@@

This is the (optional) step of saving the analyzed images.
%
The ``Output Image Directory'' field is where the analyzed images will be stored; this directory will be created if it does not exist. 
%
The images will be stored in the output directory; the name format will be the same as in the input directory.

@@

There are two parameters
\begin{description}
\item[Number of Images in Average Trajectory] This is the desired number of images in the average trajectory. 
%
\item[Width of Averaging Window] This is the (approximate) width of the averaging window. Each snapshot in the average trajectory will be the (weighted) average of approximately the width number of images. Increasing this value will result in a smoother average trajectory; decreasing this value will reveal more finer-scale detail. 
% 
\end{description}

>>>>

\newcommand{\SI}[0]{\textit{SI Materials and Methods}}

@@

The basic steps are as follows:
\begin{enumerate}
   \item Read in images
   \item Preprocess the images (smooth and/or equalize channels, mean-center, etc.)
   \item Run vector diffusion maps to register and order preprocessed images
   \item (Optional) Save registered and ordered images
   \item (Optional) Compute an average trajectory
\end{enumerate}

@@

The fields are as follows
\begin{description}
%
\item[2D/3D] Select ``2D'' if the images are traditional 2D images. Select ``3D'' if the images are z-stacks.
%
\item[Image Directory] Enter the directory where the images or z-stacks are stored. This directory can be an absolute path (e.g. \texttt{C:/Documents/Images}) or a path relative to the directory where this code is stored (e.g. \texttt{../Images})
%
\item[Image Prefix] The prefix for all image file names. The images are assumed to named as the image prefix, followed by a number. 
%
\item[Stack Prefix] (only for 3D z-stacks) The prefix for all stack folder names. Each stack is assumed to be stored in a single folder, with the folder's name as the stack prefix, followed by a number. 
%
\item[Image Extension] The extension of the images (e.g., tif, jpg)
%
\item[Number of Images] The number of images or z-stacks. Images are assumed to be stored as the image prefix, followed by a two-digit index, where the indices run from 1 to the number of images (e.g., image\_prefix01.tif, image\_prefix02.tif, etc.). For three-dimensional images, the z-stacks are assumed to be stored in individual folders whose names are the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of stacks (e.g., z\_stack\_prefix01/, z\_stack\_prefix02/, etc.)
%
\item[Number of Images in Stack] (only for z-stacks) The number of images per z-stack. Images are assumed to be stored in the individual stack directories as the stack prefix, followed by a two-digit index, where the indices run from 1 to the number of images in the stack (e.g., stack\_prefix01.tif, stack\_prefix02.tif, etc.). 
%
\end{description}

After populating the required fields, select the ``Read Images'' button to read in the images.
%
After reading the images, (optional) select the ``Show Images'' button to display the images (z-stacks will be displayed as maximum intensity projections).

@@

\begin{description}
\item[Weight] Each channel in the images is scaled by the weight factor, where all weights at 1 does not adjust the channel intensities. For grayscale images, only the first weight is used. 
%
\item[Blur] Each channel in the images is blurred using a disc factor with radius equal to a fraction of the image as indicated by the ``blur'' field. For example, setting the blur to 5\% means that the image is blurred using a disc filter with radius $0.05 \times$ the number of pixels. 
%
\item[Normalize] If the normalize option is selected for a specific channel, then Contrast-Limited Adaptive Histogram Equalization is applied to that channel to normalize the intensities across the image. This is important for signals whose absolute intensity is not meaningful or informative. 
%
\item[Use to Mean-Center] If the mean-center option is selected for any of the channels, then the object is centered in the image by detecting the edges of the object given by the selected channels. If no mean-center options are selected, then the object is not mean-centered.
%
\item[Resize Images] If the resize images option is selected, then each image is scaled/dialated so that the object occupies 80\% of the total image frame. This removes any size effects/variations in the images. 
%
\item[Number of Pixels] The number of pixels wanted in the images for analysis. This does {\em not} have to be the same number of pixels as the true resolution of the images. The images are assumed to be square. 
%
\end{description}

After populating the required fields, select the ``Apply Image'' button to apply the selected preprocessing options.
%
After preprocessing the images, (optional) select the ``Show Images'' button to display the preprocessed images (z-stacks will be displayed as maximum intensity projections).

@@

After preprocessing the images, one can apply various analysis algorithms. 
%
There are three parameters
%
\begin{description}
%
\item[Angular Discretization (degrees)] (required for image registration) This is the angular discretization used to search for the pairwise alignments between images. For example, if the angular discretization is 10$^\circ$, then the pairwise alignments are computed by computing the image discrepancy after 10$^\circ$ rotation, after a 20$^\circ$ rotation, ..., up to a 360$^\circ$ rotation, and then selecting the rotation which minimizes the discrepancy. Lower values of the angular discretization increase the accuracy of the method, but also increase the computational cost. 
%
\item[Kernel Scale (*median)] This is the scaling of the diffusion maps kernel, in multiples of the median of the pairwise distances (so that $\epsilon$ in the diffusion maps algorithm, as outlined in \SI, is this value times the median of the pairwise distances). This parameter should be $\mathcal{O}(1)$; empirically, we have found that values of $0.25-0.5$ often yield good results for our data sets. 
%
\item[Apply to...]  This determines which images are shown and/or saved as the analyzed images. Select the ``Apply to Raw Images'' option to apply the computed registration and/or ordering to the original images; select the ``Apply to Preprocessed Images'' option to apply the computed registration and/or ordering to the preprocessed images (after blurring, normalization, etc.)
%
\end{description}

There are four analysis options
\begin{description}
%
\item[Register] (this option is only for two-dimensional images) This registers the imaging data set using vector diffusion maps to compute the optimal rotations. 
%
\item[Order] This orders the imaging data set using diffusion maps. {\em Important:} This option assumes that the imaging data set is preregistered! This option is intended for imaging data sets that were preregistered using some {\em a priori} knowledge about the system. 
%
\item[Register + Order] (this option is only for two-dimensional images) This registers and orders the imaging data set using vector diffusion maps to compute the optimal rotations and the temporal ordering. 
%
\item[Register + Order z-stacks] (this option is only for three-dimensional z-stacks) This registers and orders the imaging data set. It first registers the z-stacks by computing the optimal rotations to align the (two-dimensional) maximum intensity projections using vector diffusion maps.  It then orders the data using diffusion maps on the (now registered) three-dimensional z-stacks. 
%
\end{description}

(Optional) After analysis select the ``Show Images'' button to display the registered and/or images (z-stacks will be displayed as maximum intensity projections).

@@

This is the (optional) step of saving the analyzed images.
%
The ``Output Image Directory'' is where the analyzed images will be stored; this directory will be created if it does not exist. 
%
The images will be stored in the output directory, and the name format will be the same as in the input directory.

@@

There are two parameters
\begin{description}
\item[Number of Images in Average Trajectory] This is the desired number of images in the average trajectory. 
%
\item[Width of Averaging Window] This is the (approximate) width of the averaging window. Each snapshot in the average trajectory will be the (weighted) average of approximately the width number of images (see \SI). Increasing this value will result in a smoother average trajectory, while decreasing this value will retain more finer-scale detail. 
% 
\end{description}

### 2015-02-01 image_analysis_paper_development2/manuscript.tex
updated paper

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
%
Furthermore, although our methods provide a temporal ordering for images, the provide no real-time information, neither do they provide a direction for the proposed ordering.
%
Determining which ``end'' of the trajectory is early and which is late is a post-processing task which requires {\em a priori} knowledge of the developmental dynamics. 
%
Furthermore, real-time information can only be extracted from either {\em a priori} knowledge of the rates of relevant developmental processes, or by making some assumption about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).

>>>>

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%
Such a manifold can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold. 
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

Unlike traditional statistical methods, our algorithms do not provide confidence intervals on predicted orderings, but could in principle be extracted from bootstrapping \citep{efron1979bootstrap}.
%
Furthermore, although our methods provide a temporal ordering for images, the provide no real-time information, neither do they provide a direction for the proposed ordering.
%
Determining the trajectory rate and direction are post-processing tasks which requires {\em a priori} knowledge of the developmental dynamics. 
%
Furthermore, real-time information can only be extracted from either {\em a priori} knowledge of the rates of relevant developmental processes, or by making some assumption about the temporal sampling of the data set (e.g., the data set uniformly samples some given time window).
